Designing Types for R, Empirically

### 2020-05-15 paper/main.tex
3 wip

R has one builtin notion of type that can be queried by the \k{typeof}
function. Figure~\ref{types} lists all of the builtin types that are
provided by the language. They are the possible return values of
\k{typeof}. There is no intrinsic notion of subtyping in R. But, in many
context a \k{logical} will convert to \k{integer}, and an \k{integer} will
convert to \k{double}.  Some odd conversions can occur in corner cases, such
as \k{1<"2"} holds and \k{c(1,2)[1.6]} returns the first element of the
vector, as the double is converted to an integer. R does not distinguish
between scalars and vectors (they are all vectors), so \code{typeof(5) ==}
\code{typeof(c(5)) == typeof(c(5,5))} \code{ == "double"}. Finally all
vectorized data types have a distinguished missing value denoted by
\code{NA}. The default type of \code{NA} is \k{logical}. We can see that
\code{typeof(NA)=="logical"}, but NA inhabits every type:
\code{typeof(c(1,NA)[2])=="double"}.

\begin{wrapfigure}{r}{6.1cm}
\footnotesize\begin{tabular}{l|l@{}}\hline
\multicolumn{2}{l}{\bf Vectorized data types:}  \\\hline
\k{logical}   & vector of boolean values\\
\k{integer}   & vector of 32 bit integer values\\
\k{double}    & vector of 64 bit floating points\\
\k{complex}   & vector of complex values\\
\k{character} & vector of strings values\\
\k{raw}       & vector of bytes\\
\k{list}      & vector of values of any type\\\hline
\multicolumn{2}{l}{\bf Scalar data types:}\\\hline
\k{NULL}      &  singleton null value\\
\k{S4}        &  instance of a S4 class \\
\k{closure}   &  function with its environment\\
\k{environment}& mapping from symbol to value \\\hline
\multicolumn{2}{l}{\bf Implementation data types:}\\\hline
\multicolumn{2}{l}{\k{special},
\k{builtin},
\k{symbol},
\k{pairlist},
\k{promise}}\\
\multicolumn{2}{l}{
\k{language},
\k{char},
\k{...},
\k{any},
\k{expression},
}\\
\multicolumn{2}{l}{
\k{externalprt},
\k{bytecode},
\k{weakref}}\\\hline
\end{tabular}\caption{Builtin Types}\label{types}\end{wrapfigure}

@@

Scalar data types include the distinguished \k{NULL} value, which is also of
ktype \k{NULL}, instance of classes written using the S4 object system,
closures and environments.  The implementation of R has a number of other
types that are mostly not used by user code, they are listed in
Figure~\ref{types} for reference.

>>>>

R has one builtin notion of type that can be queried by the \k{typeof}
function. Figure~\ref{rtypes} lists all of the builtin types that are
provided by the language. They are the possible return values of
\k{typeof}. There is no intrinsic notion of subtyping in R. But, in many
context a \k{logical} will convert to \k{integer}, and an \k{integer} will
convert to \k{double}.  Some odd conversions can occur in corner cases, such
as \k{1<"2"} holds and \k{c(1,2)[1.6]} returns the first element of the
vector, as the double is converted to an integer. R does not distinguish
between scalars and vectors (they are all vectors), so \code{typeof(5) ==}
\code{typeof(c(5)) == typeof(c(5,5))} \code{ == "double"}. Finally all
vectorized data types have a distinguished missing value denoted by
\code{NA}. The default type of \code{NA} is \k{logical}. We can see that
\code{typeof(NA)=="logical"}, but NA inhabits every type:
\code{typeof(c(1,NA)[2])=="double"}.

\begin{wrapfigure}{r}{6.1cm}
\footnotesize\begin{tabular}{l|l@{}}\hline
\multicolumn{2}{l}{\bf Vectorized data types:}  \\\hline
\k{logical}   & vector of boolean values\\
\k{integer}   & vector of 32 bit integer values\\
\k{double}    & vector of 64 bit floating points\\
\k{complex}   & vector of complex values\\
\k{character} & vector of strings values\\
\k{raw}       & vector of bytes\\
\k{list}      & vector of values of any type\\\hline
\multicolumn{2}{l}{\bf Scalar data types:}\\\hline
\k{NULL}      &  singleton null value\\
\k{S4}        &  instance of a S4 class \\
\k{closure}   &  function with its environment\\
\k{environment}& mapping from symbol to value \\\hline
\multicolumn{2}{l}{\bf Implementation data types:}\\\hline
\multicolumn{2}{l}{\k{special},
\k{builtin},
\k{symbol},
\k{pairlist},
\k{promise}}\\
\multicolumn{2}{l}{
\k{language},
\k{char},
\k{...},
\k{any},
\k{expression},
}\\
\multicolumn{2}{l}{
\k{externalprt},
\k{bytecode},
\k{weakref}}\\\hline
\end{tabular}\caption{Builtin Types}\label{rtypes}\end{wrapfigure}

@@

Scalar data types include the distinguished \k{NULL} value, which is also of
ktype \k{NULL}, instance of classes written using the S4 object system,
closures and environments.  The implementation of R has a number of other
types listed in Figure~\ref{rtypes} for reference.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Design A Type Language for R}%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\renewcommand{\null}{\texttt{\textbf{null}}\xspace}
\newcommand{\any}{\texttt{\textbf{any}}\xspace}
\newcommand{\environment}{\texttt{\textbf{environment}}\xspace}
\newcommand{\expression}{\texttt{\textbf{expression}}\xspace}
\newcommand{\Language}{\texttt{\textbf{language}}\xspace} %% Upper case!
\newcommand{\externalptr}{\texttt{\textbf{externalptr}}\xspace}
\renewcommand{\symbol}{\texttt{\textbf{symbol}}\xspace}
\newcommand{\pairlist}{\texttt{\textbf{pairlist}}\xspace}
\newcommand{\weakref}{\texttt{\textbf{weakref}}\xspace}
\renewcommand{\int}{\texttt{\textbf{int}}\xspace}
\newcommand{\chr}{\texttt{\textbf{chr}}\xspace}
\newcommand{\dbl}{\texttt{\textbf{dbl}}\xspace}
\newcommand{\lgl}{\texttt{\textbf{lgl}}\xspace}
\newcommand{\clx}{\texttt{\textbf{clx}}\xspace}
\newcommand{\raw}{\texttt{\textbf{raw}}\xspace}
\newcommand{\VARGS}{\texttt{\textbf{...}}\xspace}
\newcommand{\T}{\ensuremath{T}\xspace}
\renewcommand{\S}{\ensuremath{S}\xspace}
\newcommand{\V}{\ensuremath{V}\xspace}
\newcommand{\A}{\ensuremath{A}\xspace}
\newcommand{\ID}{\ensuremath{ID}\xspace}
\newcommand{\F}{\ensuremath{F}\xspace}
\newcommand{\B}{\ensuremath{B}\xspace}
\newcommand{\FUN}[2]{\ensuremath{\langle #1 \rangle \rightarrow #2}\xspace}
\newcommand{\STRUCT}[1]{\texttt{\textbf{struct}}\ensuremath{\langle #1\rangle}\xspace}
\newcommand{\LIST}[1]{\texttt{\textbf{list}}\ensuremath{\langle #1\rangle}\xspace}
\newcommand{\CLASS}[1]{\texttt{\textbf{class}}\ensuremath{\langle #1\rangle}\xspace}
\newcommand{\VEC}[1]{#1\k{[]}\xspace}
\newcommand{\NAVEC}[1]{\k{\string^}\!#1\k{[]}\xspace}

This section sets out to propose a candidate design for a type language that
can be used to describe the arguments and return values of functions in R.
The goal is not to be the final design but rather a starting point for an
iterative process.

Fig.~\ref{types} presents our type language. An early design choice was to
stay close to the R language and only depart in small, hopefully,
non-controverwsial ways.  Functions types have the form \FUN{A_1, \dots,
  A_n}\T where each $\A_i$ argument is either a type \T or \VARGS, a
variable length argument list. The rest of this section details and
motivates our design.

\begin{figure}[!h] \noindent \small  \centering\begin{minipage}{.45\linewidth}
\begin{tabular}{lclr}
\T& ::= & \null         & \it unit type\\
  & |   & \any          & \it top type\\
  & |   & \environment  & \it environment type \\
  & |   & \Language     & \it language type \\
  & |   & \symbol       & \it symbol type \\
  & |   & \S            & \it scalar type \\
  & |   & \V            & \it vector type \\
  & |   & \T~\k{|}\T    & \it union type \\
  & |   & \k{?} \T      & \it nullable type \\
  & |   & \FUN{A_1, ..., A_n}\T & \it function type \\
  & |   & \STRUCT{F_1, ..., F_n} & \it struct type \\
  & |   & \LIST{T}      & \it list type\\
  & |   & \CLASS{ID_1, ..., ID_n}& \it class type\vspace{5pt}\\
\end{tabular}\end{minipage}   \hfill
\begin{minipage}{.45\linewidth}\begin{tabular}{lclr}
\A & ::= & \T & \it arguments \\
  & |   & \VARGS \vspace{5pt} \\
\F & ::= & \ID\, :\,T & \it struct field\vspace{5pt}\\
\V & ::= & ~\VEC\S &\\
  & |   & \NAVEC\S & \it na vector types \vspace{5pt}\\
\S & ::= & \int & \\
  & |   & \chr \\
  & |   & \dbl \\
  & |   & \lgl \\
  & |   & \clx \\
  & |   & \raw \vspace{5pt}\\
\end{tabular}\end{minipage}\caption{The R type language}\label{types}
\end{figure}

\subsection{Basic Types}

The workhorse types in R are the vectorized primitives: integers, doubles,
complex numbers, logicals (booleans), characters, and raws (bytes).  These
are vectorized in that they are always considered to be vectors by the R
runtime: for instance, even ``scalar'' numbers are considered to be
unit-length vectors to R.  These primitive vectors are homogeneous, in that
a vector of doubles contains only doubles.

\subsubsection{Possibly-NA Primitives}

scalars and often do high-level
dimensionality checks on their data.  To illustrate, consider the code in
Figure~\ref{fig:scalar-vector-example}.

%
\subsection{Classes}

R has more than one notion of type.  The classes of function arguments are
used by R as part of its dispatch mechanism, the most prevalent of which
(\AT{we have that number, too!}) is S3 dispatch, in which R dispatches on
the (S3) class of the first argument to a function, selecting the
appropriate method to call for the class.

Classes are very prevalent in R \AT{number}, and so we include a
\code{class<class-name>} ... \AT{todo} \typetracer was outfit to report the
class of values, but some care needed to be taken so as to not overreport
class types, as {\it all} values have a class in R.  The main classes we
were interested in were matrix, array (higher dimensional matrices), and
data.frames, in addition to any user-defined classes.  In many cases
\AT{number?}, values with a user-defined class are in reality lists with
named elements, or structs.  Including class types, and having \typetracer
report them in place of structs, significantly reduced the noise in the
analysis and allowed us to gather more focused data.

When set to ignore classes, \typetracer reports \AT{type} as the type of the
function, as compared with \AT{other type} when configured to report
classes.
%
A value can have multiple classes in R.  This is common practice in the
\code{dplyr} package, which extends the functionality of data frames: a
\code{data.frame} that passes through \code{dplyr} functions gains two
additional classes (\code{tbl} and \code{tbl_df}), bringing the total to at
least 3, in addition to whichever other classes that data frame happened to
have.  Note that when looking to perform S3 dispatch, R will try to match
{\it any} of the classes of the first argument to a dispatched function, so
each of these classes are important.

%
\subsection{Nullable Types}
